---
title: "HW 1"
author: "Claire Chen"
date: today
format: pdf
editor: visual
---

```{r message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(haven)
library(knitr)
library(report)
library(broom)

df <- read_sav("https://github.com/felixthoemmes/hwdatasets/blob/master/hw1.sav?raw=true")
```

# Question 1

Check the type of the variables in the data set and change the "group" variable as a factor.

```{r}
df

df <- df |>
  mutate(
    group = factor(group))
```

Report the mean and the standard deviation of the reaction time variable in the overall sample.

```{r}
sample_summary <- df |>
  summarize(mean = mean(reaction), sd = sd(reaction))

sample_summary |>
  kable(digits = 2,
        caption = "Descriptive Statistics of Reaction Time in Overall Sample")
```

Report the mean and the standard deviation of the reaction times but grouped by the variable “group”.

```{r}
group_summary <- df |>
  group_by(group) |>
  summarize(mean = mean(reaction), sd = sd(reaction))
  
group_summary |>
  kable(digits = 2,
        caption = "Descriptive Statistics of Reaction Time in Groups 0 and 1")
```

Report all of your results in a well-formatted table.

```{r}
combined_summary <- bind_rows(group_summary, sample_summary)

combined_summary |>
  kable(digits = 2,
        caption = "Descriptive Statistics of Overall and Groups' Reaction Time")
```

# Question 2

Form a new variable “logdeviationreaction”.

```{r}
df <- df |>
  mutate(logdeviationreaction = log(abs(df$reaction-mean(df$reaction))))
```

Report the mean, and standard deviation of variable “logdeviationreaction”.

```{r}
df |>
  summarize(mean = mean(logdeviationreaction), sd = sd(logdeviationreaction)) |>
  kable(digits = 2,
        caption = "Descriptive Statistics of Variable 'logdeviationreaction'")
```

# Question 3

Report the mean and standard deviation of “logdeviationreaction” variable for individuals assigned to group 0.

```{r}
df |>
  filter(group == 0) |>
  summarize(mean = mean(logdeviationreaction), sd = sd(logdeviationreaction)) |>
  kable(digits = 2,
        caption = "Decriptive Statistics of Group 0 legdeviationreaction")
```

# Question 4

Form z-scores of the previously created reaction time variable "logdeviationreaction" and save them in a new column, called “zreaction”.

```{r}
df <- df |>
  mutate(zreaction = (logdeviationreaction-mean(logdeviationreaction))
         /sd(logdeviationreaction))
```

Print the first 10 cases of zreaction only in your document.

```{r}
head(df$zreaction, 10) |>
  kable(caption = "First 10 Cases of 'zreaction'")
```

# Question 5

Filter all cases that have a “zreaction” time that is larger than \|1\|, and then count cases that fulfill this criterion within each group defined by the “group” variable. Use summarize() and the n() function to obtain your results and print them in your document.

```{r}
df |>
  filter(abs(zreaction) >= 1) |>
  group_by(group) |>
  summarize(cases = n()) |>
  kable(digits = 2,
        caption = "Numbers of 'zreaction' Larger than |1| in Groups 0 and 1")
```

# Question 6

Create a new data set, where you select only the reaction time variable, and then filter only those cases in which reaction time was larger than the median reaction time in the whole sample. Then print all cases of this new data set.

```{r}
new_df <- df |>
  select(reaction) |>
  filter(reaction > median(reaction))

new_df |>
  kable(caption = "Cases with Reaction Time Larger Than Median Reaction Time")
```

# Question 7

Use the “report_sample” function to obtain descriptive statistics for the reaction time variable grouped by the “group” variable. Print the results of the report_sample function in your report.

```{r}
df |>
  select(reaction, group) |>
  group_by(group) |>
  report_sample()
```
